---
title: "EPL PPG over the last 10 years"
author: "Joseph Pym"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
---

# Welcome

This is a very simple R code to see the Points Per Game achieved by each team in the English Football League over the past 10 years, inspired by a [tweet](https://twitter.com/LawrenceTanner1/status/1393240204438347780) I saw about AFC Wimbledon's record. First, we need to load the data, which we do from [football-data.co.uk](http://www.football-data.co.uk). 

```{r loaddata}
library(dplyr)

co.we <- c("E0", "E1", "E2", "E3")  #Website country codes (football-data)
seasons <- c("1112", "1213", "1314", "1415", "1516", "1617", "1718", "1819", "1920", "2021")
matchdataTemp <- NULL; matchdata <- NULL
for (i in seasons){
  for (j in 1:4){
    matchdataTemp <- read.csv(paste0('https://www.football-data.co.uk/mmz4281/', i, '/', co.we[j],'.csv'), fileEncoding = 'latin1')
    matchdataTemp <- matchdataTemp[ ,c("Div", "Date", "HomeTeam", "AwayTeam", "FTHG", "FTAG", "FTR")]
    matchdata <- rbind(matchdata, matchdataTemp)
  }
}
matchdata <- na.omit(matchdata)
```

# Point Data

Next, we need to find the point data, which we do by first using the Unique tool from the dplyr package to find all unique team names and creating a new dataframe with these. For each row in the PointData dataframe, we compare against each row in the MatchData dataframe: this code checks if the Home Team or Away Team is the team of interest, and if so, allocates points (3 for a win, 1 for a draw).

```{r pointdata}
pointdata <- NULL
for (i in unique(matchdata$HomeTeam)){
  pointdata <- rbind(pointdata, i)
}
pointdata <- data.frame(pointdata)
pointdata$points <- with(pointdata, 0)
pointdata$games <- with(pointdata, 0)
colnames(pointdata) <- c("team", "points", "games")
rownames(pointdata) <- 1:nrow(pointdata)

for (i in 1:nrow(pointdata)){
  for (j in 1:nrow(matchdata)){
    if (matchdata$HomeTeam[j] == pointdata$team[i]){
      if (matchdata$FTR[j] == "H"){pointdata$points[i] <- pointdata$points[i] + 3}
      else if (matchdata$FTR[j] == "D"){pointdata$points[i] <- pointdata$points[i] + 1}
      pointdata$games[i] <- pointdata$games[i] + 1}
    
    else if (matchdata$AwayTeam[j] == pointdata$team[i]){
      if (matchdata$FTR[j] == "A"){pointdata$points[i] <- pointdata$points[i] + 3}
      else if (matchdata$FTR[j] == "D"){pointdata$points[i] <- pointdata$points[i] + 1}
      pointdata$games[i] <- pointdata$games[i] + 1
    }
  }
}
```

# Final Table

Finally, we find each teams Points Per Game.

```{r ppg}
pointdata$ppg <- with(pointdata, 0)
for (i in 1:nrow(pointdata)){
  pointdata$ppg[i] <- pointdata$points[i] / pointdata$games[i]
}
pointdata[order(pointdata$ppg),]
```











