\documentclass[9pt,a4paper]{article}
\usepackage[a4paper, total={6.5in, 10in}]{geometry}
\usepackage{amsmath} \usepackage{amssymb} \usepackage{wrapfig}
\usepackage{fancyvrb} \usepackage{appendix} \usepackage{graphicx}\usepackage{url}
\usepackage{caption,subcaption} \usepackage{hyperref}
\pagenumbering{Roman}
\title{\scshape{Curriculum Vitae}\\Joseph Pym}\author{josephpym@outlook.com\\Mobile: 07894 526468}\date{Apartment 32, Caister Hall, Coventry, CV1 5PE\\\texttt{www.linkedin.com/in/josephpym} \\ \texttt{jep00.github.io}}

\begin{document}
\maketitle\setcounter{secnumdepth}{0}
\begin{center}\rule{0.5\textwidth}{0.3pt}\end{center}
\begin{center} % Introduction 
\noindent Final year BSc Mathematics \& Statistics student at Coventry University, with excellent mathematical, statistical, time management and logical skills, as well as an enthusiastic team worker with strong attention to detail. Passionate about statistics, data science and artificial intelligence; highly motivated with experience in various statistical computing programs; always keen to learn new skills.\end{center}
\begin{center}\rule{0.5\textwidth}{0.3pt}\end{center}

\section{SKILLS}
\noindent Proficient using R, with experience using Python, MATLAB, Minitab, SPSS, Simulink.\\
Highly skilled on \LaTeX, Microsoft Office and iWork.\\
Clean UK driving license.
\section{EDUCATION}
\textsc{Coventry University}: 2018---Present (Exp. Graduation 2021)\\
\hangindent=0.7cm{ Final year \textbf{BSc Mathematics and Statistics} student;\\
	On course for a \textbf{First Class} degree (First/second year average $>$ 80\%);\\
	Modules have included Further Calculus \& Algebra; Statistical Computing; Statistical Design \& Modelling; Advanced Topics in Statistics; Operational Research Techniques; Mathematical Modelling; Linear Statistical Models; Coding \& Cryptography; Optimisation.\\

\noindent\textsc{Hutton Grammar Sixth Form}: 2016---2018\\
\hangindent=0.7cm A \textemdash\ Mathematics and Statistics;\\A \textemdash\ Chemistry;\\C \textemdash\ Physics.\\Also recieved an A the Extended Project Qualification.\\

\noindent\textsc{Hutton CofE Grammar School}: 2011---2016\\
\hangindent=0.7cm 13 GCSEs: 3 A*'s (incl. Mathematics), 6 A's (incl. English Language and Literature) and 4 B's.
\section{OTHER QUALIFICATIONS}
\begin{tabular}{p{0.45\textwidth}p{0.45\textwidth}}
	The Duke of Edinburgh's Award & Bronze \\
	BTEC Aviation Studies & Pass
\end{tabular}


\pagebreak
\section{PROJECTS}
\noindent\textsc{Third Year University Project:} {``Assessing the Accuracy of Betting Odds in European Football''}\\
\hangindent=0.7cm \emph{{Ongoing}}\\
Dissertation and presentation on determining whether or not bookmakers offer accurate odds for football matches, using a range of statistical tools and methods for analysis. Coding done using R, with a \LaTeX\ write-up, dealing with over 75,000 matches. Created an algorithm to place bets, with a comparison against many runs of random betting.\vspace{3mm}

\noindent\textsc{First Year University Assignment:} MATLAB Simulation \& Algebra Coursework\\
\hangindent=0.7cm \textsc{grade: }\textbf{90\%}\\
Designed a simulation in which a random number of balls are generated within plotted `walls', moving at random and bouncing off the walls, with added features such as a trace behind the balls, `gravity', energy (and loss of); and the use of symbolic algebra tools to differentiate, integrate, and plot functions using MATLAB. The write-up was completed in \LaTeX. This was the first large coding project I undertook, and it taught me a lot about planning, writing and correcting code: skills I have used greatly since. \vspace{3mm}

\noindent\textsc{EPQ:} {``Will there be Supersonic Transport again by the year 2030, and if so, will it be sustainable?''\\
\hangindent=0.7cm\textsc{grade: }\textbf{A} ($>$90\%)\\Independently researched 5,000 word report followed by a presentation of my findings, adhering to strict timelines; researching a relatively unknown topic; and critically evaluated all sources.\\From this, a lot was learnt about dealing with deadlines and time management.

\section{WORK EXPERIENCE}
\noindent\textsc{Southport \& Ainsdale Golf Club}---Clubhouse Waiting Staff.\\
\hangindent=0.7cm 2015\textemdash Present\\Member of the waiting-on and bar staff at a two-time Ryder Cup host and regular championship venue, working on major club events, internal and external functions and day-to-day operations in the clubhouse and halfway house.\\My role has taught me skills such as financial handling (cashing up tills), time management (setting up function rooms in time), teamwork, dealing with stressful and busy times, and leadership (when running the pass). 
\section{HOBBIES \& INTERESTS}
\noindent\textsc{Cricket}\\
\hangindent=0.7cm Represented various different teams at multiple levels, including $1^{\textrm{st}}$ XI cricket; captaining Hutton Grammar School $2^{\textrm{nd}}$ XI and New Longton CC u19 in the National Under 19 Blast; elected to captain Coventry University CC (CUCC) $2^{\textrm{nd}}$ XI.\\
During second year, was elected CUCC secretary, with roles including organising events and arranging fixtures against other university sides, and the successful running of our \emph{Covid Cricket} charity league throughout lockdown caused by Covid-19, raising over £800.\\
Elected member of the Sport Coventry committee, looking after sport as a whole at the University, involved in key decisions and major event organisation (such as Varsity).\\
Current vice-president of CUCC, helping the running of the club. Roles performed include completing a report into the club's social media; helping run the club's Movember efforts; the recruitment of first-year students throughout the Covid-19 pandemic.\vspace{5mm}

\noindent\textsc{Air Training Corps.}\\
\hangindent=0.7cm Member of 967 (BAe Warton) and 471 (Tarleton \& Hesketh Bank) Air Cadet squadrons from 2013-2018, reaching the rank of Cadet Sergeant. Awarded Cadet of the Year (2016); earnt BTEC Aviation Studies and Bronze DofE; qualified as an Instructor Cadet; attended Junior and Senior NCO training camps, adventure sports camp and annual camp on an active RAF station; ran the squadron social media and website.
\end{document}